Early Warning Signals and the Prosecutor's Fallacy

### 2012-06-05 fallacy.tex
eliminating words

  Early warning signals have been proposed to forecast the possibility
  of a critical transition, such as the eutrophication of a lake,
  the collapse of a coral reef, or the end
  of a glacial period.  Because such transitions often occur on scales that
  can be difficult to approach by experimental manipulation, research
  has often relied on historical observations as a source of natural experiments.
  Here we examine a critical difference between selecting systems for
  study based on the fact that we have observed a critical transition
  and those systems for which we wish to forecast the approach of a
  transition. This difference arises by conditionally selecting systems
  known to experience a transition of some sort and failing to account
  for this bias this introduces -- a statistical error often known as the
  Prosecutor's Fallacy.  By analyzing simulated systems that have
  experienced transitions purely by chance, we reveal an elevated rate
  of false postives in common warning signal statistics. We further
  demonstrate a model-based approach which is less subject to this bias
  than these more commonly used summary statistics.
  In using historical examples of collapse to test and validate methods
  for early warning signals, care must be taken to avoid this fallacy,
  either through the use methods less sensitive to this effect or ideally
  by greater reliance on Experimental studies with replicates avoid this 
  pitfall entirely.  More such studies are needed to provide data suitable
  for evaluating methods for early warning.

@@

\begin{equation}
  P(E|I) = P(I|E) \frac{P(E)}{P(I)}
\end{equation} }

With little other evidence supporting the conviction, The California
Supreme Court reversed the decision, and the case became a widely
recognized example of the Prosecutor's Fallacy~\citep{Thompson1987}.
While DNA evidence has brought a rise in the number of probability-based
convictions in the courtroom, and a corresponding rise in the frequency
with which suspects are falsely accused, trial juries are not the
only ones who are susceptible to misinterpreting probabilistic evidence.
In this paper we explore how a similar misconception can arise from the
use of historical data to evaluate methods for detecting early warning
signals of critical transitions.

@@

Not all sudden transitions are caused by some guilty process slowly
driving the system over a tipping point, that is, the kind of process
that early warning signals are designed to detect.  Some systems may
experience such transitions purely by chance, leaving a stable state on
an extremely unlikely excursion that happens to stray to far from the
stable attractor~\citep[\emph{e.g.} ][consider this possibility in 
transitions that arise from analyzing historical climate record]{Ditlevsen2010, Lenton2011}. 
Like the evidence presented before the California Supreme
Court in 1968, the chance of observing such an ``innocent'' transition 
a priori may be very small, but when selected from a historical record
of many possible transitions, this possibility can no longer be ignored.

In principle, as such chance transitions do not involve the shrinking
eigenvalue that drives the critical slowing down and increasing variance
phenomena upon which several early warning signal statistics are based,
such data should not appear to display a warning signal. If this is the case,
then we can use historical data with more confidence when seeking to 
test early warning patterns, knowing that we will not mistake such examples
as cases of the bifurcation process we seek to detect.

Figure 1 shows a schematic illustrating critical transitions under
each of these scenarios.  In the left panel, the system experiences 
a bifurcation that forces it over the edge -- just the kind of data 
we would want to test an early warning signal.  In the right panel,
a similar-looking trajectory emerges from a simulation of a stable 
system.  While the simulation of the bifurcation scenario shown on
the left produces a similar transition every time, the transition 
shown on the right is somewhat less likely, occurring in only 1\% of
simulations.

@@

\section{Methods and Results}
We approach the problem by simulating a system that will not undergo the
saddle node bifurcation that underlies the usual early warning signal
indicators.  This simulation approach allows us to determine whether
examining historical events is a good way to test the utility of these
indicators.  We simulated 20,000 replicates of a stochastic individual-based
birth-death process with an Allee threshold~\citep{Courchamp2008}, which
arises from positive fitness effects at low densities.  Above the Allee
threshold the population returns to a positive equilibrium size, while
below the threshold the population decreases to zero in expectation. The
simulation uses the Gillespie algorithm to provide an exact implementation
of the master equation of the birth-death process,

\begin{align}
  \frac{dP(n,t)}{dt} &= b_{n-1} P(n-1,t) + d_{n+1}P(n+1,t) - (b_n+d_n) P(n,t)  \label{master} \\
    b_n &= \frac{e K n^2}{n^2 + h^2} \\
    d_n &= e n + a
\end{align}
a model with a linear death rate and density dependent birth rate
that drives the Allee effect at low densities and limits growth
at high densities.  The key feature of this model is the alternate
stable states introduced by this effect; other functional forms for
Eq.~\eqref{master} could serve equally well for these simulations~\citep[see
\emph{e.g.}][]{Scheffer2001}.  Though this system can be forced
through a saddle-node bifurcation by increasing the death rate, in
these simulations all parameters are held constant and no bifurcation
occurs.  Consequently we do not anticipate an early warning signal of
an approaching bifurcation.  The simulation starts from the positive
equilibrium population size.  Though the chance of a transition across
the Allee threshold in any given time step are small, given enough
time this system will eventually experience such a rare event driving
the population extinct.  We evolved each replicate 50,000 time steps,
sampling the system every 50 time steps.  In this time window 266 of the
1,000 replicates experience population collapse.  To keep the examples of
comparable sample size, we focus on a section of the data 500 time points
prior to the system approaching the transition.

@@

To test for the presence of a warning signal in these indicators we
computed values of Kendall's $\tau$ for both indicators for each of
the 266 replicates.  Kendall's $\tau$ takes values in $(-1, 1)$, and is
frequently used to identify an increasing trend ($\tau > 0 )$ in early
warning signals.~\citep{Dakos2008, Dakos2011}.  The distribution
of $\tau$ values observed across these replicates is shown in
Figure~\ref{fig:indicator}.  More positive values indicate stronger
trends. The common statistical significance test used for Kendall's $\tau$
assumes independence of the variables to achieve a normal distribution,
this isn't satisfied for indicators coming from a time series and
computed with a sliding window.  For a more robust comparison, we
compare the distribution of this statistic from all the simulations to
the distribution conditioned on experiencing a chance transition to the
alternative stable state, that is, from the replicates which are known
to crash.  To avoid an effect of sample size the time series are all chosen
to be the same length.  As these examples have been selected from a system
that is not approaching a bifurcation, an indicator that does not suffer
from the bias of Prosecutor's fallacy should have the same distribution
in both cases.   Annotated, parallelized code to replicate the analysis
and produce the figures is provided in the appendix.  (Currently at
\href{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}).

@@

For each of these replicates we also take a model-based approach, estimating 
parameters for an approximate linear model of the system approaching a
saddle node bifurcation.

\begin{equation}
  dX = \sqrt{ r_t } (\phi(r_t) - X_t) dt +
  \sigma\sqrt{\phi(r_t) } d B_t \label{LSN}
\end{equation}
where \( \phi(r_t) = \sqrt{r_t} +\theta \) is the equilibrium position.  
This model emerges from the canonical form of the saddle-node bifurcation,

\begin{equation}
  \frac{dX}{dt} = r - x^2
\end{equation}
where we introduce an explicit parameter $\theta$ to account for rescaling
to the units of the data,

\begin{equation}
   \frac{dX}{dt} = r - (x-\theta)^2
\end{equation}
and then consider the corresponding stochastic process where the noise is
proportional to the square root of the mean, as is the case for demographic
noise processes,

\begin{equation}
  \frac{dX}{dt} = r - (x-\theta)^2 + \sqrt{ \beta (r + (x-\theta)^2 } dB_t
\end{equation}
$r$ remains the bifurcation parameter which we imagine changing with time,
$r_t$, then linearizing the above model in neighborhood of
\(\hat x = \sqrt{r_t} +\theta =: \phi(r_t)\), we recover Equation~\eqref{LSN}.
These models can be both be fit to the data by maximum likelihood estimation
based on the equation for the evolution of their moments, as explained
in~\citet{Boettiger2012b}.  Though this approximation is more general than 
the original Markov process (as other models containing a saddle-node bifurcation
share the same approximation), it is also easier to estimate reliably from 
the data since it describes only the process near the equilibrium where the
data is available.  We assume that $r_t$ is varying at constant rate, 
according to the model

\begin{equation}
 r_t = m t + r_0 
\end{equation}
and estimate $m$, $r_0$, $\theta$, and $\beta$ by maximum likelihood on each
of the simulated datasets.  Estimates of $m < 0 $ are expected in systems 
approaching a bifurcation, while for stable systems $m$ should be approximately zero.

Of the 266 simulations, only six had estimates with machine precision different
from zero, the greatest magnitude among these was $-7.1 \times 10^{-12}$,
which would still correspond to a negligible change in $r_0$ over the 50,000 time steps
of the simulation.  Consequently, the model-based estimation shows no 
evidence of bias on data that has been selected conditional on collapse.

@@

Historical examples of systems that have undergone sudden transitions seem
to offer an invaluable opportunity to test methods for the detection of
early warning signals that could have foreseen the collapse.  However,
such data must be used with care, as the process of selecting systems
already known to have collapsed opens up the possibility that these
transitions were the result of rare chance events rather than examples
of the more deterministic tipping point behaviour we wish to detect.
When we think about forecasting potential collapse in a single system,
these purely-chance transitions may be rare enough to ignore.  When we
comb the historical record for any available data on sudden transitions,
our odds of finding these chance transitions may be much higher.

>>>>

  Early warning signals have been proposed to forecast the possibility
  of a critical transition, such as the eutrophication of a lake,
  the collapse of a coral reef, or the end
  of a glacial period.  Because such transitions often unfold on temporal and spatial scales that
  can be difficult to approach by experimental manipulation, research
  has often relied on historical observations as a source of natural experiments.
  Here we examine a critical difference between selecting systems for
  study based on the fact that we have observed a critical transition
  and those systems for which we wish to forecast the approach of a
  transition. This difference arises by conditionally selecting systems
  known to experience a transition of some sort and failing to account
  for the bias this introduces -- a statistical error often known as the
  Prosecutor's Fallacy.  By analyzing simulated systems that have
  experienced transitions purely by chance, we reveal an elevated rate
  of false postives in common warning signal statistics. We further
  demonstrate a model-based approach that is less subject to this bias
  than these more commonly used summary statistics.
  In using historical examples of collapse to test and validate methods
  for early warning signals, care must be taken to avoid this fallacy,
  either through the use of methods less sensitive to this effect or ideally
  by greater reliance on experimental studies with replicates avoid this 
  pitfall entirely.  More such studies are needed to provide data suitable
  for evaluating methods for early warning.

@@

\begin{equation}
  P(E|I) = P(I|E) \frac{P(E)}{P(I)}
\end{equation}

} With little other evidence supporting the conviction, The California
Supreme Court reversed the decision, and the case became a widely
recognized example of the Prosecutor's Fallacy~\citep{Thompson1987}.
Trial juries are not the
only ones susceptible to misinterpreting probabilistic evidence.
In this paper we explore how a similar misconception can arise from the
use of historical data to evaluate methods for detecting early warning
signals of critical transitions.

@@

Not all sudden transitions are caused by some ``guilty'' process slowly
driving the system over a tipping point -- the kind of process
that early warning signals are designed to detect.  Some systems may
experience such transitions purely by chance, leaving a stable state on
an extremely unlikely excursion that happens to stray to far from the
stable attractor~\citep[\emph{e.g.} ][consider this possibility in 
transitions that arise from analyzing historical climate record]{Ditlevsen2010, Lenton2011}. 
Like the evidence presented before the California Supreme
Court in 1968, the chance of observing such an ``innocent'' transition 
a priori may be very small, but when selected from a historical record
of many possible transitions, this possibility can no longer be ignored.

Figure 1 shows a schematic illustrating critical transitions under
each of these scenarios.  In the left panel, the system experiences a
bifurcation  and should contain an early warning signal.  In the right
panel, a similar-looking trajectory emerges from a simulation of a stable
system which should not contain a warning signal.  While the simulation of
the bifurcation scenario shown on the left produces a similar transition
every time, the transition shown on the right is somewhat less likely,
occurring in only 1\% of simulations.

@@

\section{Methods and Results}
To investigate if early warning signals are vulnerable to this fallacy,
we simulating a system that is not driven towards a bifurcation (such as
in the right-hand panel).  This simulation approach allows us to determine whether
examining historical events is a good way to test the utility of these
indicators.  We simulated 20,000 replicates of a stochastic individual-based
birth-death process with an Allee threshold~\citep{Courchamp2008}, which
arises from positive fitness effects at low densities.  Above the Allee
threshold the population returns to a positive equilibrium size, whereas
below the threshold the population decreases to zero. The
simulation uses the Gillespie algorithm to provide an exact implementation
of the master equation of the birth-death process,

\begin{align}
  \frac{dP(n,t)}{dt} &= b_{n-1} P(n-1,t) + d_{n+1}P(n+1,t) - (b_n+d_n) P(n,t)  \label{master} \\
    b_n &= \frac{e K n^2}{n^2 + h^2} \\
    d_n &= e n + a
\end{align}

a model with a linear death rate and density-dependent birth
rate that drives the Allee effect at low densities and limits
growth at high densities.  The key feature of this model is the
alternate stable states introduced by this effect; other functional
forms for Eq.~\eqref{master} could serve equally well for these
simulations~\citep[see \emph{e.g.}][]{Scheffer2001}.  Though this
system can be forced through a bifurcation by increasing
the death rate, in these simulations all parameters are held constant
and no bifurcation occurs.  Consequently we do not anticipate an early
warning signal of an approaching bifurcation.  The simulation starts
from the positive equilibrium population size.  Though the chance of
a transition across the Allee threshold in any given time step is 
small, given enough time this system will eventually experience such a
rare event driving the population extinct.  We ran each replicate over
50,000 time unitss, sampling the system every 50 time units.  In this
time window 266 of the 1,000 replicates experience population collapse.
To keep the examples of comparable sample size, we focus on a section of
the data 500 time points prior to the system approaching the transition.

@@

To test for the presence of a warning signal in these indicators we
computed values of Kendall's $\tau$ for both indicators for each of
the 266 replicates.  Kendall's $\tau$ takes values in $(-1, 1)$, and is
frequently used to identify an increasing trend ($\tau > 0 )$ in early
warning signals~\citep{Dakos2008, Dakos2011}.  The distribution
of $\tau$ values observed across these replicates is shown in
Figure~\ref{fig:indicator}.   
%The common statistical significance test used for Kendall's $\tau$
%assumes independence of the variables to achieve a normal distribution,
%this isn't satisfied for indicators coming from a time series and
%computed with a sliding window.  For a more robust comparison, 
We compare the distribution of $\tau$ from all the simulations to
the distribution conditioned on experiencing a chance transition to the
alternative stable state.  To avoid an effect of sample size the time series are all chosen
to be the same length.  
%As these examples have been selected from a system
%that is not approaching a bifurcation, an indicator that does not suffer
%from the bias of Prosecutor's fallacy should have the same distribution
%in both cases.   
Annotated, parallelized code to replicate the analysis
and produce the figures is provided in the appendix.  (Currently at
\href{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}{https://github.com/cboettig/earlywarning/blob/prosecutor/inst/examples/fallacy.md}).

@@

For each of these replicates we also take a model-based
approach, estimating parameters for an approximate linear
model of the system approaching a saddle node bifurcation, as
described~\citet{Boettiger2012b}.  In this model, the parameter $m$
describes the approach towards the saddle-node bifurcation.  Estimates
$m < 0 $ are expected in systems approaching a bifurcation, while for
stable systems $m$ should be approximately zero. None of the estimates
across the 266 simulations differed from zero in our study, hence the
model-based estimation shows no evidence of bias on data that has been
selected conditional on collapse.

@@

Historical examples of systems that have undergone sudden transitions 
offer a unique opportunity to test methods for the detection of
early warning signals that could have foreseen the collapse.  However,
such data must be used with care, as the process of selecting systems
already known to have collapsed opens up the possibility that these
transitions were the result of rare chance events rather than examples
of the more deterministic tipping point behaviour we wish to detect.
When we think about forecasting potential collapse in a single system,
these purely-chance transitions may be rare enough to ignore.  When we
comb the historical record for any available data on sudden transitions,
our odds of finding these chance transitions may be much higher.
